Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-28 paper/IFL_2016_reviews_RESPONSE.txt

We thank the reviews for in-depth reviews. For convenience, here is a
brief description of how we addressed the reviewers' recommendations:

@@

2) The treatment of related work must be significantly improved.

Added section 2: Related work.

3) A clear and concise overview of the constructs of the language has
to be added, along with a good characterisation of at least the
semantics of the core constructs "sample" and "observe", as formal
as possible.

Rewrote section 5. Language, with the syntax of Clojure forms
supported, core library support, and section 5.3 on
probabilistic forms, with details on their syntax and semantics, 
related to (new) section 4: Probability of a problem execution.

@@

4) Clarify why the language constructs justify building a new language
instead of making a library.

Added a general clarification at the end of section 3, design outline, 
and an additional explanation at the end of subsection 5.3.

@@

7) Make it clear that section addressing of checkpoints refers to a known
addressing scheme, and reduce amount of detail if necessary to save
space, referring to earlier work instead.

Upon discussing this point with the reviewers we discovered that
there was misunderstanding, and addressing of checkpoints is
indeed a contribution of the paper:

> The reviewers have re-read this part, and it is now clear there was
> a misunderstanding on their part.
> 
> Thus, ignore point 7 as it stands, and instead just make it clear
> both up-front and where the addressing scheme is discussed that this
> is novel and one of the central contributions of the paper.
> 
> Additionally, in general, try to bring out the central contributions
> and ideas in a way that does not get obscured by too many details
> specific to your setting. I believe many people who seek to implement
> similar languages can learn a lot from your work, and it would be
> great if that is made as easy as possible!

Added addressing of checkpoints to the list of contributions in
the beginning of the paper. In subsection 8.3, Addressing of
checkpoint, changed the language 'uses a different scheme' to
'introduces a new scheme'.

@@

8) Bring the convincing example at page 9 significantly forward.

Case study is now Section 6, right after language definition and
before discussing any implementation issues.

9) Substantiate performance claims by at least some concrete results
from benchmarking.

Added section 10 Performance, with performance comparison
between Anglican and Clojure, as well as between Anglican and
other probabilistic programming environments, with quantitative evaluation 
on a set of probabilistic models.

### 2017-03-05 paper/paper.tex
edits to clarify the different meanings of checkpoint

A probabilistic program, or query, mostly runs deterministic code,
except for \texttt{sample} and \texttt{observe} forms, which are
probabilistic in nature. For this reason, Anglican can be implemented
as a regular programming language, except for the handling of these
two special forms, which we refer to as checkpoints. At these
checkpoints, Anglican programs must allow the inference algorithm to
step in, recording information and affecting control flow. This can
be implemented through coroutines/cooperative multitasking, and
parallel execution/preemptive multitasking, as well as through
explicit maintenance of program continuations at checkpoints.
Anglican follows the latter option.

@@

The purpose of probabilistic forms \texttt{sample} and
\texttt{observe} is to interrupt deterministic computation and
transfer control to the inference algorithm.  Practically, this
is achieved by  returning a \textbf{checkpoint} instead of
invoking a continuation in a CPS-transformed program.  
Checkpoints are represented by values of the
corresponding record types
(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe})
in Clojure. The types contain fields specific to each form, as well as the
continuation; calling the continuation resumes the computation.
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist v) 'cont)
(->observe dist v cont $state)
\end{lstlisting}
Here \texttt{->sample} and \texttt{->observe} in the
CPS-transformed expressions are constructors for Clojure
records, and they take values of their fields as arguments.

>>>>

A probabilistic program, or query, mostly runs deterministic code.
Aside from the special forms \texttt{sample} and \texttt{observe},
which are probabilistic in nature, Anglican can be implemented as a
regular programming language. At \texttt{sample} and
\texttt{observe} forms, normal deterministic execution is 
interrupted and Anglican programs must allow the inference algorithm
to step in, recording information and affecting control flow. We refer these points in the execution as checkpoints. Handling of 
checkpoints can be implemented through coroutines/cooperative multitasking, and parallel execution/preemptive multitasking, as well as through explicit maintenance of program continuations. Anglican follows the latter option.

@@

The purpose of probabilistic forms \texttt{sample} and
\texttt{observe} is to interrupt deterministic computation and
transfer control to the inference algorithm.  Practically, this
is achieved by returning a record that represents the state of program execution at the checkpoint, rather than invoking a continuation. The Clojure record used to interrupt execution at 
\texttt{sample} and \texttt{observe} forms are \texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} and \texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe} respectively. These records contain a unique identifier (which we will discuss in detail in section \ref{sec:addressing}), the arguments to the special form, the continuation, and the state.
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample 'S1 dist cont $state)
=> (cps-of-expression '(observe dist v) 'cont)
(->observe 'O1 dist v cont $state)
\end{lstlisting}
Here \texttt{->sample} and \texttt{->observe} in the
CPS-transformed expressions are constructors for Clojure
records, and they take values of their fields as arguments.
Calling the continuation resumes the computation.
